---
title: "Assignment 2: Spatial Analysis and Visualization"
subtitle: "Healthcare Access and Equity in Pennsylvania"
author: Jun Luu
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Assignment Overview

**Learning Objectives:** - Apply spatial operations to answer
policy-relevant research questions - Integrate census demographic data
with spatial analysis - Create publication-quality visualizations and
maps - Work with spatial data from multiple sources - Communicate
findings effectively for policy audiences

------------------------------------------------------------------------

## Part 1: Healthcare Access for Vulnerable Populations

### Research Question

**Which Pennsylvania counties have the highest proportion of vulnerable
populations (elderly + low-income) living far from hospitals?**

Your analysis should identify counties that should be priorities for
healthcare investment and policy intervention.

### Required Analysis Steps

Complete the following analysis, documenting each step with code and
brief explanations:

#### Step 1: Data Collection (5 points)

Load the required spatial data: - Pennsylvania county boundaries -
Pennsylvania hospitals (from lecture data) - Pennsylvania census tracts

**Your Task:**

```{r}
#| include: false
# Load required packages
library(sf)
library(tidyverse)
library(tigris)
library(tidycensus)
library(scales)
library(patchwork)
library(here)
library(knitr)
library(kableExtra)

# Load spatial data
pa_counties <- st_read("data/Pennsylvania_County_Boundaries.shp")
pa_tracts <- tracts(state = "PA", cb = TRUE)
pa_hospitals <- st_read("data/hospitals.geojson")

# Check that all data loaded correctly
glimpse(pa_counties)
glimpse(pa_tracts)
glimpse(pa_hospitals)

# 1. How many hospitals are in your dataset? 
num_hospitals <- nrow(pa_hospitals)
cat("Number of hospitals:", num_hospitals, "\n")

# 2. How many census tracts? 
num_tracts <- nrow(pa_tracts)
cat("Number of census tracts:", num_tracts, "\n")

# 3. What coordinate reference system is each dataset in? 
cat("County CRS:", st_crs(pa_counties)$input, "\n")
cat("Tract CRS:", st_crs(pa_tracts)$input, "\n")
cat("Hospital CRS:", st_crs(pa_hospitals)$input, "\n")

options(tigris_progress = FALSE)
pa_tracts <- st_transform(pa_tracts, st_crs(pa_counties))

```

**Questions to answer:** - How many hospitals are in your dataset? - How
many census tracts? - What coordinate reference system is each dataset
in?

------------------------------------------------------------------------

#### Step 2: Get Demographic Data

Use `tidycensus` to download tract-level demographic data for
Pennsylvania.

**Required variables:** - Total population - Median household income -
Population 65 years and over (you may need to sum multiple age
categories)

**Your Task:**
```{r}
#| include: false
acs_vars <- load_variables(2022, "acs5", cache = TRUE)

# Get demographic data from ACS

census_api_key(Sys.getenv("368bf145f527c34904bbbc75ef3158887059279a"))
```

```{r}
acs_vars <- load_variables(2022, "acs5", cache = TRUE)

# Get demographic data from ACS

census_api_key(Sys.getenv("368bf145f527c34904bbbc75ef3158887059279a"))

variables <- c(
  median_household_income = "B19013_001",
  total_population = "B01003_001",
  male_65_75 = "B01001A_014",
  male_75_85 = "B01001A_015",
  male_85_over = "B01001A_016",
  female_65_75 = "B01001A_029",
  female_75_85 = "B01001A_030",
  female_85_over = "B01001A_031"
)

pa_tract_acs <- get_acs(
  geography = "tract",
  state = "PA",
  variables = variables,
  year = 2022,
  survey = "acs5",
  cache_table = TRUE, 
  output = "wide"
)

pa_tract_acs <- pa_tract_acs %>%
  mutate(
    pop_65_overE = male_65_75E + male_75_85E + male_85_overE +
                   female_65_75E + female_75_85E + female_85_overE
  )

# Join to tract boundaries
tracts_with_data <- pa_tracts %>%
  left_join(pa_tract_acs, by = "GEOID")

# - What year of ACS data are you using? 
acs_year <- 2022
cat("ACS Year:", acs_year, "\n")

# How many tracts have missing income data? 
missing_income <- sum(is.na(tracts_with_data$median_household_incomeE))
cat("Number of tracts with missing income data:", missing_income, "\n")

# What is the median income across all PA census tracts? 
median_income_statewide <- median(tracts_with_data$median_household_incomeE, na.rm = TRUE)
cat("Median income across all PA tracts: $", round(median_income_statewide, 0), "\n")

```

**Questions to answer:** - What year of ACS data are you using? - How
many tracts have missing income data? - What is the median income across
all PA census tracts?

------------------------------------------------------------------------

#### Step 3: Define Vulnerable Populations

Identify census tracts with vulnerable populations based on TWO
criteria: 1. Low median household income (choose an appropriate
threshold) 2. Significant elderly population (choose an appropriate
threshold)

**Your Task:**

```{r}
# Filter for vulnerable tracts based on your criteria
tracts_with_data <- tracts_with_data %>%
  mutate(
    percent_65_over = (pop_65_overE / total_populationE) * 100
  )

income_threshold <- quantile(tracts_with_data$median_household_incomeE, probs = 0.25, na.rm = TRUE)
age_threshold <- 20

tracts_with_data <- tracts_with_data %>%
  mutate(
    low_income = median_household_incomeE <= income_threshold, 
    elderly = percent_65_over >= age_threshold,    
    vulnerable = low_income & elderly 
  )

# 1. What income threshold did you choose and why?
cat("Income threshold: $", income_threshold, 
    ", which is the bottom quartile of income in Pennsylvania\n")

# 2. What elderly population threshold did you choose and why?
cat("Elderly population threshold:", age_threshold, 
    "% , based on the Census Bureau. \n")

# 3. How many tracts meet your vulnerability criteria?
num_vulnerable <- sum(tracts_with_data$vulnerable, na.rm = TRUE)
cat("Number of vulnerable tracts:", num_vulnerable, "\n")

# 4. What percentage of PA census tracts are considered vulnerable?
total_tracts <- nrow(tracts_with_data)
pct_vulnerable <- (num_vulnerable / total_tracts) * 100
cat("Percentage of vulnerable tracts in PA:", round(pct_vulnerable, 2), "%\n")


```

**Questions to answer:** - What income threshold did you choose and
why? - What elderly population threshold did you choose and why? - How
many tracts meet your vulnerability criteria?\
- What percentage of PA census tracts are considered vulnerable by your
definition?

------------------------------------------------------------------------

#### Step 4: Calculate Distance to Hospitals

For each vulnerable tract, calculate the distance to the nearest
hospital.

**Your Task:**

```{r}
# Transform to appropriate projected CRS
pa_counties_proj <- st_transform(pa_counties, 3365)
pa_hospitals_proj <- st_transform(pa_hospitals, 3365)
pa_tracts_proj <- st_transform(tracts_with_data, 3365)

# Calculate distance from each tract centroid to nearest hospital
all_centroids <- st_centroid(pa_tracts_proj)

all_dist <- st_distance(all_centroids, pa_hospitals_proj)
all_centroids$nearest_hosp_ft <- apply(all_dist, 1, min)

all_centroids <- all_centroids %>%
  mutate(nearest_hosp_mi = nearest_hosp_ft / 5280)

pa_centroids <- pa_tracts_proj %>%
  left_join(
    all_centroids %>%
      st_drop_geometry() %>%
      select(GEOID, nearest_hosp_mi),
    by = "GEOID"
  )

num_vul = sum(pa_centroids$vulnerable, na.rm = TRUE)

vulnerable_tracts <- pa_centroids %>%
  st_drop_geometry() %>%
  filter(vulnerable == TRUE)

# 1. What is the average distance to the nearest hospital for vulnerable tracts?
avg_dist <- mean(vulnerable_tracts$nearest_hosp_mi, na.rm = TRUE)
cat("Average distance to nearest hospital:", round(avg_dist, 2), "miles\n")

# 2. What is the maximum distance to the nearest hospital?
max_dist <- max(vulnerable_tracts$nearest_hosp_mi, na.rm = TRUE)
cat("Maximum distance to nearest hospital:", round(max_dist, 2), "miles\n")

# 3. How many vulnerable tracts are more than 15 miles from the nearest hospital?
num_far <- sum(vulnerable_tracts$nearest_hosp_mi > 15, na.rm = TRUE)
cat("Number of vulnerable tracts > 15 miles from nearest hospital:", num_far, "\n")

# 4. What percentage of vulnerable tracts are >15 miles away?
over_15mi <- (num_far / num_vul) * 100
cat("Percentage of vulnerable tracts > 15 miles from nearest hospital:", round(over_15mi, 2), "%\n")

```

**Requirements:** - Use an appropriate projected coordinate system for
Pennsylvania - Calculate distances in miles - Explain why you chose your
projection: I used EPSG: 26918 (UTM Zone 18N) as the projected
coordinate system because it preserves distance accuracy across
Pennsylvania.

**Questions to answer:** - What is the average distance to the nearest
hospital for vulnerable tracts? - What is the maximum distance? - How
many vulnerable tracts are more than 15 miles from the nearest hospital?

------------------------------------------------------------------------

#### Step 5: Identify Underserved Areas

Define "underserved" as vulnerable tracts that are more than 15 miles
from the nearest hospital.

**Your Task:**

```{r}
# Create underserved variable
underserved_tracts <- pa_centroids %>%
  st_drop_geometry() %>%
  filter(vulnerable == TRUE & nearest_hosp_mi > 15)

pa_centroids <- pa_centroids%>%
  mutate(
    underserved = vulnerable & nearest_hosp_mi > 15
  )

num_underserved = sum(pa_centroids$underserved == TRUE, na.rm = TRUE)
# 1. How many tracts are underserved?
cat("Number of underserved tracts:", num_underserved, "\n")

# 2. What percentage of vulnerable tracts are underserved?
pct_underserved <- (num_underserved / num_vul) * 100
cat("Percentage of vulnerable tracts that are underserved:", round(pct_underserved, 2), "%\n")

# 3. Does this surprise you? Why or why not?
cat("This honestly does surprise me, because I thought that the number would be higher. I think that I could have decreased the elderly population percentage to 15%, but I will continue to use the information determined by the Census Bureau.\n")


```

**Questions to answer:** - How many tracts are underserved? - What
percentage of vulnerable tracts are underserved? - Does this surprise
you? Why or why not?

------------------------------------------------------------------------

#### Step 6: Aggregate to County Level

Use spatial joins and aggregation to calculate county-level statistics
about vulnerable populations and hospital access.

**Your Task:**

```{r}
# Spatial join tracts to counties
county_tract_proj <- pa_counties_proj %>%
  st_join(pa_centroids)


# Aggregate statistics by county
county_summary <- county_tract_proj %>%
  st_drop_geometry() %>%
  group_by(COUNTY_NAM) %>%
  summarise(
    num_vul_tracts = sum(vulnerable, na.rm = TRUE),
    num_und_tracts = sum(vulnerable & nearest_hosp_mi > 15, na.rm = TRUE),
    pct_und_tracts = round((num_und_tracts/num_vul_tracts)*100, 2),
    avg_vul_dist = round(mean(nearest_hosp_mi[vulnerable], na.rm = TRUE), 2),
    total_vul_pop = round(sum(total_populationE[vulnerable], na.rm = TRUE), 2)
  )

# 1. Which 5 counties have the highest percentage of underserved vulnerable tracts?
top5_pct_underserved <- county_summary %>%
  slice_max(pct_und_tracts, n = 5)

kable(top5_pct_underserved, caption = "Top 5 counties with the highest percentage of underserved vulnerable tracts")

# 2. Which counties have the most vulnerable people living far from hospitals?
top_5_und_counties <- county_summary %>%
  arrange(desc(total_vul_pop)) %>%
  slice_head(n = 5)

kable(top_5_und_counties, caption = "Counties with the most vulnerable people living far from hospitals")

# 3. Are there any patterns in where underserved counties are located?
county_summary_map <- pa_counties_proj %>%
  left_join(county_summary, by = "COUNTY_NAM")

# Quick visualization
ggplot(county_summary_map) +
  geom_sf(aes(fill = pct_und_tracts)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey90") +
  labs(
    title = "Percentage of Underserved Vulnerable Tracts by County",
    fill = "% Underserved"
  ) +
  theme_minimal()

```

**Required county-level statistics:** - Number of vulnerable tracts -
Number of underserved tracts\
- Percentage of vulnerable tracts that are underserved - Average
distance to nearest hospital for vulnerable tracts - Total vulnerable
population

**Questions to answer:** - Which 5 counties have the highest percentage
of underserved vulnerable tracts? - Which counties have the most
vulnerable people living far from hospitals? - Are there any patterns in
where underserved counties are located?

------------------------------------------------------------------------

#### Step 7: Create Summary Table

Create a professional table showing the top 10 priority counties for
healthcare investment.

**Your Task:**

```{r}
# Create and format priority counties table
county_priority <- county_summary %>%
  mutate(
    priority_score = round((.5*scale(pct_und_tracts))+(.5*scale(total_vul_pop)), 2),
  ) %>%
  arrange(desc(priority_score)) %>%
  slice_head(n = 10)

county_priority %>%
  select(
    County = COUNTY_NAM,
    `Vulnerable Tracts` = num_vul_tracts,
    `Underserved Tracts` = num_und_tracts,
    `Underserved (%)` = pct_und_tracts,
    `Avg Distance (mi)` = avg_vul_dist,
    `Vulnerable Population` = total_vul_pop,
    `Priority Score` = priority_score
  ) %>%
  kable(
    caption = "Top 10 Priority Counties for Healthcare Investment in Pennsylvania<br>(Based on Vulnerable Populations and Hospital Access)",
    format = "html",
    align = c("l", "r", "r", "r", "r", "r")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 13
  ) %>%
  row_spec(0, bold = TRUE, background = "#002855", color = "white")
  

```

**Requirements:** - Use `knitr::kable()` or similar for formatting -
Include descriptive column names - Format numbers appropriately (commas
for population, percentages, etc.) - Add an informative caption - Sort
by priority (you decide the metric)

------------------------------------------------------------------------

## Part 2: Comprehensive Visualization

Using the skills from Week 3 (Data Visualization), create
publication-quality maps and charts.

### Map 1: County-Level Choropleth

Create a choropleth map showing healthcare access challenges at the
county level.

**Your Task:**

```{r}
# Create county-level access map
pa_counties_summary <- pa_counties_proj %>%
  left_join(county_summary, by = "COUNTY_NAM")

ggplot() +
  geom_sf(
    data = pa_counties_summary,
    aes(fill = pct_und_tracts),
    color = "white",
    size = 0.2
  ) +
  geom_sf(
    data = pa_hospitals_proj,
    fill = "red",
    color = "white",
    size = 1.2,
    shape = 21,
    stroke = 0.3,
    alpha = 1
  ) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "% Underserved\n(Vulnerable Tracts)",
    labels = function(x) paste0(x, "%"),
    na.value = "grey90"
  ) +
  labs(
    title = "Healthcare Access Challenges in Pennsylvania by County",
    subtitle = "Counties shaded by the % of underserved tracts"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10)),
    plot.caption = element_text(size = 9, color = "gray40", hjust = 0),
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 9),
    legend.position = "right"
  )

```

**Requirements:** - Fill counties by percentage of vulnerable tracts
that are underserved - Include hospital locations as points - Use an
appropriate color scheme - Include clear title, subtitle, and caption -
Use `theme_void()` or similar clean theme - Add a legend with formatted
labels

------------------------------------------------------------------------

### Map 2: Detailed Vulnerability Map

Create a map highlighting underserved vulnerable tracts.

**Your Task:**

```{r}
# Create detailed tract-level map

ggplot() +
  geom_sf(
    data = pa_counties_proj,
    fill = "grey95",
    color = "white",
    size = 0.3
  ) +
  geom_sf(
    data = pa_centroids,
    aes(geometry = geometry),
    color = "grey85",
    size = 0.1,
    alpha = 0.4
  ) +
  geom_sf(
    data = pa_centroids %>% filter(underserved == TRUE),
    aes(geometry = geometry),
    fill = "blue",
    color = "white",
    size = 0.15,
    alpha = 0.8
  ) +
  geom_sf(
    data = pa_hospitals_proj,
    fill = "red",
    color = "white",
    size = 1.2,
    shape = 21,
    stroke = 0.3,
    alpha = 1
  ) +
  labs(
    title = "Underserved Vulnerable Census Tracts in Pennsylvania",
    subtitle = "Tracts with vulnerable populations located more than 15 miles from the nearest hospital"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12, margin = margin(b = 10)),
    plot.caption = element_text(size = 9, color = "gray40", hjust = 0),
    legend.position = "none"
  ) +
  coord_sf(expand = FALSE)

```

**Requirements:** - Show underserved vulnerable tracts in a contrasting
color - Include county boundaries for context - Show hospital
locations - Use appropriate visual hierarchy (what should stand out?) -
Include informative title and subtitle

------------------------------------------------------------------------

### Chart: Distribution Analysis

Create a visualization showing the distribution of distances to
hospitals for vulnerable populations.

**Your Task:**

```{r}
# Create distribution visualization
ggplot(pa_centroids %>% filter(vulnerable), 
       aes(x = nearest_hosp_mi, y = total_populationE)) +
  geom_point(alpha = 0.6, color = "blue", size = 1.2) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Distance to Nearest Hospital vs. Vulnerable Population Size",
    x = "Distance to Nearest Hospital (mi)",
    y = "Vulnerable Population per Tract",
    caption = "Each point represents a vulnerable census tract in Pennsylvania. Tracts farther from hospitals\nwith larger vulnerable populations indicate high-priority areas for healthcare investment."

  ) +
  theme_minimal(base_size = 12)


```

**Suggested chart types:** - Histogram or density plot of distances -
Box plot comparing distances across regions - Bar chart of underserved
tracts by county - Scatter plot of distance vs. vulnerable population
size

**Requirements:** - Clear axes labels with units - Appropriate title -
Professional formatting - Brief interpretation (1-2 sentences as a
caption or in text)

------------------------------------------------------------------------

## Part 3: Bring Your Own Data Analysis

Choose your own additional spatial dataset and conduct a supplementary
analysis.

### Challenge Options

Choose ONE of the following challenge exercises, or propose your own
research question using OpenDataPhilly data
(https://opendataphilly.org/datasets/).

**Note these are just loose suggestions to spark ideas - follow or make
your own as the data permits and as your ideas evolve. This analysis
should include bringing in your own dataset, ensuring the projection/CRS
of your layers align and are appropriate for the analysis (not lat/long
or geodetic coordinate systems). The analysis portion should include
some combination of spatial and attribute operations to answer a
relatively straightforward question**

------------------------------------------------------------------------

#### Education & Youth Services

**Option A: Educational Desert Analysis** - **Data:** Schools,
Libraries, Recreation Centers, Census tracts (child population) -
**Question:** "Which neighborhoods lack adequate educational
infrastructure for children?" - **Operations:** Buffer schools/libraries
(0.5 mile walking distance), identify coverage gaps, overlay with child
population density - **Policy relevance:** School district planning,
library placement, after-school program siting

**Option B: School Safety Zones** - **Data:** Schools, Crime Incidents,
Bike Network - **Question:** "Are school zones safe for walking/biking,
or are they crime hotspots?" - **Operations:** Buffer schools (1000ft
safety zone), spatial join with crime incidents, assess bike
infrastructure coverage - **Policy relevance:** Safe Routes to School
programs, crossing guard placement

------------------------------------------------------------------------

#### Environmental Justice

**Option C: Green Space Equity** - **Data:** Parks, Street Trees, Census
tracts (race/income demographics) - **Question:** "Do low-income and
minority neighborhoods have equitable access to green space?" -
**Operations:** Buffer parks (10-minute walk = 0.5 mile), calculate tree
canopy or park acreage per capita, compare by demographics - **Policy
relevance:** Climate resilience, environmental justice, urban forestry
investment ---

#### Public Safety & Justice

**Option D: Crime & Community Resources** - **Data:** Crime Incidents,
Recreation Centers, Libraries, Street Lights - **Question:** "Are
high-crime areas underserved by community resources?" - **Operations:**
Aggregate crime counts to census tracts or neighborhoods, count
community resources per area, spatial correlation analysis - **Policy
relevance:** Community investment, violence prevention strategies ---

#### Infrastructure & Services

**Option E: Polling Place Accessibility** - **Data:** Polling Places,
SEPTA stops, Census tracts (elderly population, disability rates) -
**Question:** "Are polling places accessible for elderly and disabled
voters?" - **Operations:** Buffer polling places and transit stops,
identify vulnerable populations, find areas lacking access - **Policy
relevance:** Voting rights, election infrastructure, ADA compliance

------------------------------------------------------------------------

#### Health & Wellness

**Option F: Recreation & Population Health** - **Data:** Recreation
Centers, Playgrounds, Parks, Census tracts (demographics) -
**Question:** "Is lack of recreation access associated with vulnerable
populations?" - **Operations:** Calculate recreation facilities per
capita by neighborhood, buffer facilities for walking access, overlay
with demographic indicators - **Policy relevance:** Public health
investment, recreation programming, obesity prevention

------------------------------------------------------------------------

#### Emergency Services

**Option G: EMS Response Coverage** - **Data:** Fire Stations, EMS
stations, Population density, High-rise buildings - **Question:** "Are
population-dense areas adequately covered by emergency services?" -
**Operations:** Create service area buffers (5-minute drive = \~2
miles), assess population coverage, identify gaps in high-density
areas - **Policy relevance:** Emergency preparedness, station siting
decisions

------------------------------------------------------------------------

#### Arts & Culture

**Option H: Cultural Asset Distribution** - **Data:** Public Art,
Museums, Historic sites/markers, Neighborhoods - **Question:** "Do all
neighborhoods have equitable access to cultural amenities?" -
**Operations:** Count cultural assets per neighborhood, normalize by
population, compare distribution across demographic groups - **Policy
relevance:** Cultural equity, tourism, quality of life, neighborhood
identity

------------------------------------------------------------------------

### Data Sources

**OpenDataPhilly:** https://opendataphilly.org/datasets/ - Most datasets
available as GeoJSON, Shapefile, or CSV with coordinates - Always check
the Metadata for a data dictionary of the fields.

**Additional Sources:** - **Pennsylvania Open Data:**
https://data.pa.gov/ - **Census Bureau (via tidycensus):** Demographics,
economic indicators, commute patterns - **TIGER/Line (via tigris):**
Geographic boundaries

### Recommended Starting Points

**If you're feeling confident:** Choose an advanced challenge with
multiple data layers. **If you are a beginner, choose something more
manageable that helps you understand the basics**

**If you have a different idea:** Propose your own question! Just make
sure: - You can access the spatial data - You can perform at least 2
spatial operations

### Your Analysis

**Your Task:**

1.  **Find and load additional data**
    -   Document your data source
    -   Check and standardize the CRS
    -   Provide basic summary statistics

```{r}
# Load your additional dataset
recreation_centers <- st_read("https://opendata.arcgis.com/api/v3/datasets/9eb26a787a6e448ba426eea7f9f0d93a_0/downloads/data?format=geojson&spatialRefId=4326")
#check crs
st_crs(pa_centroids)
st_crs(recreation_centers)
#only use philly data
phila_centroids <- pa_centroids %>%
  filter(COUNTYFP == "101")
#transform crs
recreation_centers <- st_transform(recreation_centers, st_crs(phila_centroids))
summary(recreation_centers)
```

**Questions to answer:** - What dataset did you choose and why? I chose
the recreation center location data to look at the question, "Is lack of
recreation access associated with vulnerable populations?" - What is the
data source and date? The data source is OpenDataPhilly and the date
created is March 2, 2022. - How many features does it contain? It
contains 11 features. - What CRS is it in? Did you need to transform it?
It is in WGS 84 and I did need to transform it.

------------------------------------------------------------------------

2.  **Pose a research question**

Write a clear research statement that your analysis will answer."Is lack
of recreation access associated with vulnerable populations?"

**Examples:** - "Do vulnerable tracts have adequate public transit
access to hospitals?" - "Are EMS stations appropriately located near
vulnerable populations?" - "Do areas with low vehicle access have worse
hospital access?"

------------------------------------------------------------------------

3.  **Conduct spatial analysis**

Use at least TWO spatial operations to answer your research question.

**Required operations (choose 2+):** - Buffers - Spatial joins - Spatial
filtering with predicates - Distance calculations - Intersections or
unions - Point-in-polygon aggregation

**Your Task:**

```{r}
# Your spatial analysis
rec_buffer <- st_buffer(recreation_centers, dist = 2640) #.5 miles
#centroids of recreation centers
phila_centroids <- phila_centroids %>%
  mutate(
    within_recreation_buffer = lengths(st_intersects(geometry, rec_buffer)) > 0
  )
#seeing results for both vulnerable and not vulnerable 
rec_summary <- phila_centroids %>%
  st_drop_geometry() %>%
  summarise(
    total_tracts = n(),
    tracts_with_access = sum(within_recreation_buffer, na.rm = TRUE),
    total_vulnerable_tracts = sum(vulnerable, na.rm = TRUE),
    vulnerable_with_access = sum(vulnerable & within_recreation_buffer, na.rm = TRUE),
    vulnerable_without_access = sum(vulnerable & !within_recreation_buffer, na.rm = TRUE),
    pct_vulnerable_with_access = round(vulnerable_with_access / total_vulnerable_tracts * 100, 1)
  )

rec_summary

```

```{r}
#to see vulnerable on ggplot
phila_centroids <- phila_centroids %>%
  mutate(
    vul_access_cat = case_when(
      !vulnerable ~ "Not vulnerable",
      vulnerable & within_recreation_buffer ~ "Vulnerable w/ access",
      vulnerable & !within_recreation_buffer ~ "Vulnerable w/o access"
    )
  )
```

```{r}
ggplot() +
  # Plot tracts colored by vulnerability + access category
  geom_sf(
    data = phila_centroids,
    aes(fill = vul_access_cat),
    color = "white",
    size = 0.1
  ) +
  
  # Recreation center buffer used
  geom_sf(
    data = rec_buffer,
    fill = "blue",
    alpha = 0.2,
    color = NA
  ) +
  
  # Recreation centers
  geom_sf(
    data = recreation_centers,
    color = "blue",
    size = 1,
    shape = 21,
    fill = "white",
    stroke = 0.6
  ) +
  
  # County boundary outline
  geom_sf(
    data = st_union(phila_centroids),
    fill = NA,
    color = "gray",
    linewidth = 0.6
  ) +
  
  # colors for each category
  scale_fill_manual(
    values = c(
      "Not vulnerable" = "gray",
      "Vulnerable w/o access" = "red",
      "Vulnerable w/ access" = "green"
    ),
    name = "Tract Category"
  ) +
  
  labs(
    title = "Recreation Access for Vulnerable Populations in Philadelphia",
  ) +
  
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    plot.caption = element_text(size = 9, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )
```

**Analysis requirements:** - Clear code comments explaining each step -
Appropriate CRS transformations - Summary statistics or counts - At
least one map showing your findings - Brief interpretation of results
(3-5 sentences)

**Your interpretation:**

I hadn't taken a look at the vulnerable tracts in Philadelphia
specifically until part 3, which was interesting. There were a lot less vulnerable tracts in Philadelphia than I had originally assumed. Reviewing the recreation center locations, there are no centers in the vulnerable tracts. This could be due to lack of funding in these vulnerable tracts. 

## Finally - A few comments about your incorporation of feedback!

Take a few moments to clean up your markdown document and then write a
line or two or three about how you may have incorporated feedback that
you recieved after your first assignment.

I hid the Census API key and data-loading code in a setup chunk, so it runs without cluttering the document, and instead I printed the key outputs (like tables and summaries) in the console or in formatted tables. This made the final report cleaner, easier to read, and more professional while keeping the analysis fully reproducible.

------------------------------------------------------------------------

## Submission Requirements

**What to submit:**

1.  **Rendered HTML document posted to your course portfolio** with all
    code, outputs, maps, and text
    -   Use `embed-resources: true` in YAML so it's a single file
    -   All code should run without errors
    -   All maps and charts should display correctly

**File naming:** `LastName_FirstName_Assignment2.html` and
`LastName_FirstName_Assignment2.qmd`

------------------------------------------------------------------------
